---
title: "Predictive modeling of claims status"
author: 'Lucas Childs, Akhil'
date: today
---

### Abstract

For the primary task, we developed a Convolutional Neural Network (CNN) for binary webpage classification using the claims-raw.RData training dataset containing 2,165 labeled webpages. The model predicts whether webpages contain "Relevant claim content" or "N/A: No relevant content." The trained CNN achieved 83.4% classification accuracy on the test set of `claims-raw.RData`, and was deployed to generate predictions on the unlabeled `claims-test.RData` dataset (929 webpages). Predicted results were saved to `preds-group[10].RData`. 

### Preprocessing

The preprocessing pipeline extracted all text content from HTML using `get_text()` from the BeautifulSoup library. This parses the HTML structure and retrieves text from paragraphs, headers, and other elements. Text cleaning involved removing URLs, email addresses, punctuation, and numeric characters, followed by splitting text strings by whitespace and converting text to lowercase. For quantitative representation, the cleaned text was converted into fixed-length sequences of integers using a vocabulary-based approach where a vocabulary of the 10,000 most frequent words was built from the training data. Each word was mapped to a unique integer index, and documents were converted to sequences of these indices, truncated or padded to a maximum length of 500 tokens to ensure uniform input dimensions for the neural network.

### Methods

The binary classification model uses a Convolutional Neural Network architecture specifically designed for text classification. The model specification includes an embedding layer (128 dimensions) to convert word indices into dense vectors, followed by three parallel 1D convolutional layers with filter sizes of 3, 4, and 5 (100 filters each) to capture n-gram patterns of different lengths. Max-pooling was used to extract the most important features and a fully connected layer with dropout (0.5) was used for regularization. Hyperparameters were selected based on common practices for text CNNs, with a vocabulary size of 10,000 words and maximum sequence length of 500 tokens. The model was trained using the Adam optimizer (learning rate 0.001) with binary cross-entropy loss over 20 epochs (not training beyond this amount because of time limitations), employing early stopping (patience=15) based on validation accuracy to prevent overfitting, with the best-performing model saved for deployment.

### Results

For the binary classification approach, we started by implementing a deep feedforward neural network, but chose a CNN as the final model for its increased classification accuracy. 

**Binary Classification: CNN metrics:**

```{r, echo=FALSE}
CNN_metrics <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity"),
  Value = c(0.833718, 0.864078, 0.806167)
)

knitr::kable(CNN_metrics)
```

## Predictive modeling of claims status (Primary Task 1B)

### Abstract

For Primary Task 1B, I developed a multi-class prediction model that assigns webpages to one of five possible claim-related categories. The workflow involved preprocessing raw HTML text, generating semantic embeddings using a SentenceTransformer model, and training a neural network classifier in PyTorch. The model achieved a validation accuracy of about 93 percent. After validation, I applied the model to the unlabeled test set and saved the predictions in `mclass_pred.RData` for integration into the team's final submission.

### Preprocessing

The raw dataset provides HTML strings under the `text_tmp` field. I used BeautifulSoup in Python to remove scripts, styles, and all HTML markup, leaving only readable text. The text was converted to lowercase, and URLs and noninformative symbols were removed. Whitespace was normalized and the cleaned version was stored as `text_clean`. This preprocessing was applied to both the training dataset and the unlabeled claim-test dataset. I saved the cleaned versions as `claims_clean_processed.csv` and `claims_test_processed.csv`.

### Methods

The modeling pipeline consisted of two main parts. First, I used the `all-MiniLM-L6-v2` SentenceTransformer model to convert each webpage into a 384-dimensional semantic embedding. These embeddings capture contextual meaning better than bag-of-words or simple token encodings.

Second, I trained a fully connected neural network in PyTorch with two hidden layers, ReLU activations, and dropout for regularization. I split the cleaned data into training and validation sets using an 80–20 stratified split. The network was trained for ten epochs using cross entropy loss and the Adam optimizer. After training and validation, I retrained the model on the full dataset to produce the deployable version used for inference.

### Results

The model performed well on the validation set, achieving approximately 93 percent accuracy. Precision, recall, and F1 scores were strong across all five classes, particularly for the more clearly defined categories such as Possible Fatality and Potentially unlawful activity. The following table summarizes the validation metrics.

```{r echo=FALSE, message=FALSE}
library(knitr)

results <- data.frame(
  Class = c("N/A: No relevant content",
            "Other claim content",
            "Physical Activity",
            "Possible Fatality",
            "Potentially unlawful activity"),
  Precision = c(0.97, 0.85, 0.85, 0.92, 0.92),
  Recall = c(0.89, 0.85, 0.98, 0.98, 0.99),
  F1 = c(0.93, 0.85, 0.91, 0.95, 0.95)
)

kable(results, caption = "Validation performance by class")
```

After validating the model, I generated predictions for all 929 unlabeled webpages in `claims-test.RData`. The predictions were saved as `mclass_pred.RData` and linked by `.id` to the group’s binary predictions from Primary Task 1A. These merged results form the final file necessary for Deliverable 3.

### Deliverables

All required files for Primary Task 1B have been included. The deployable model (`mclass_torch_model.pt`) and label mapping (`mclass_label_mapping.json`) are stored in the `results/` directory. The preprocessing, training, and prediction scripts are stored in `scripts/`, including a demonstration script showing how to load the saved model and run inference on new text. The final predictions for the test set were saved to `preds_group10.RData` and provided to the group for integration into the final submission.


